\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{booktabs}
\usepackage{hyperref}

\title{CSCI 2590 Homework 4 -- Part 1 Report}
\author{Yvonne Yu (yy5919)}
\date{\today}

\begin{document}
\maketitle

\section*{Repository Link (Required)}
GitHub repository for code (Part I and Part II):
\href{https://github.com/Yvonne-Yu217/CS2590_hw_3_yy}{https://github.com/Yvonne-Yu217/CS2590\_hw\_3\_yy}

\section*{Q2.1: Transformation Design and Rationale}
I use a \textbf{token-level stochastic transformation} that combines synonym replacement with a typo fallback:
\begin{itemize}
	\item For each alphabetic token, with probability $p = 0.15$, attempt to transform it.
	\item First, query WordNet synonyms for the token and collect candidates different from the original form.
	\item If synonyms exist, replace the token with one randomly selected synonym.
	\item If no synonym exists and token length is greater than 1, create a typo by swapping one character with a neighboring character.
	\item Non-alphabetic tokens are left unchanged; all transformed tokens are detokenized back to a sentence.
\end{itemize}

\textbf{Why this is reasonable.}
This perturbation models common real-world noise: lexical paraphrases (users may choose different words with similar meanings) and keyboard errors. In most cases, sentiment polarity should remain unchanged, so this is a realistic out-of-distribution evaluation while still preserving task semantics.

\section*{Q3.1: Data Augmentation Report and Analysis}
I trained two systems:
\begin{itemize}
	\item \textbf{Base}: trained on original IMDB training split.
	\item \textbf{Augmented}: trained on original training split plus 5,000 transformed examples.
\end{itemize}

\subsection*{Results}
\begin{table}[h]
\centering
\begin{tabular}{lcc}
\toprule
Model & Original Test Accuracy & Transformed Test Accuracy \\
\midrule
Base & 0.93448 & 0.90840 \\
Augmented & 0.93432 & 0.92108 \\
\bottomrule
\end{tabular}
\caption{Part 1 evaluation results.}
\end{table}

\subsection*{Analysis}
\textbf{(1) Did transformed-set performance improve after augmentation?} Yes.
Transformed accuracy increases from 0.90840 to 0.92108 ($+0.01268$), showing improved robustness to lexical/typo perturbations.

\textbf{(2) How did augmentation affect original-set performance?}
Original accuracy changes from 0.93448 to 0.93432 (difference $-0.00016$), which is negligible. So augmentation improved robustness without meaningfully hurting in-distribution performance.

\textbf{Intuitive explanation.}
By exposing the model to transformed variants during training, the classifier learns more stable features that are less sensitive to surface-form differences (synonyms and minor misspellings). This improves invariance under realistic perturbations while retaining useful sentiment cues from the original data.

\textbf{One limitation.}
This augmentation is local and lexical: it does not capture deeper distribution shifts such as discourse changes, sarcasm, domain transfer, or major syntactic rephrasings. Therefore, gains on this specific transformed set may not fully translate to broader OOD scenarios.

\section*{Notes on Required Output Files}
The corresponding generated files for Part 1 submission are:
\begin{itemize}
	\item \texttt{out\_original.txt}
	\item \texttt{out\_transformed.txt}
	\item \texttt{out\_augmented\_original.txt}
	\item \texttt{out\_augmented\_transformed.txt}
\end{itemize}

\section*{AI Usage Disclosure}
I used AI assistance for debugging environment/runtime issues (dependency compatibility and dataset loading), command-line workflow support on HPC, and report-language polishing. I reviewed and validated all code and experimental results before inclusion.

\end{document}
